Leadership across cultures and contexts: an exploratory analysis

### 2020-05-15 leadership_across_cultures_contexts.Rmd
Working on introduction

Research on leadership within anthropology extends to the birth of the discipline. Early on, anthropologists (and some sociologists) realized that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Anthropologists have documented the emergence of leaders among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and in diverse contexts, such as kin groups [@dussart2000politics], ritual [@singh_cultural_2017], work groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

@@

Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and nonindustrial societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Such approaches can also incorporate the anthropological insight that diversity of cultural, social, and ecological contexts are strong drivers of human behavior and group dynamics [@wiessner_risk_1982; @henrich_big_2015].

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group contex. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

<!-- GROUP CONTEXT and cost PROBLEMS -->
Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking subtleties in the nature of leadership and followership across social contexts and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [for discussions on costs of leadership see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014]. Outstanding questions for evolutionary leadership scholars include the degree to which dimensions of leadership are associated with between-group population-level variation and within-group variation in social contexts, as well as identifying the associated benefits *and* costs both leaders and followers incur.
<!-- CAN PULL SOME COST CITATIONS FROM NHB PAPER -->

<!-- THIS PARAGRAPH CAN BE INCORPORATED ABOVE ????? -->
Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with socially sanctioned rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

<!-- CLOSING -->
We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across nonindustrial non-WEIRD human societies (we include detiled review of related literature on WEIRD populations in the SI). We then outline the aims and methods of the current cross-cultural study -- which focuses on exploring patterns among non-WEIRD populations -- before presenting results and interpretations. Through novel exploratory analyses encompassing a broad range of cultural diversity and social contexts we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

## The functions of leadership across cultures and contexts
<!-- INTRO -->
Despite the universality of leadership, the functions leaders serve across populations and social contexts are highly diverse and expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when group demands require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].

Theoretical literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. In the absence of institutionalized organizational structure leaders often manage within-group coordination and collective action, including conflict resolution within or between households or kin groups, managing individual investments to group projects, and administering sanctions when individuals fail to appropriately contribute or seriously violate social norms. Demands of between-group interactions, including both conflict and cooperation are also common across societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Representing the group and managing internal and external relationships and conflicts are expected to be the most salient functions of smaller-scale community leaders. Within communities leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are suggested to be fairly widespread functions of leaders.

@glowacki_leadership_2015 draw on empirical and ethnographic results in suggesting institutionalized leadership roles among small-scale societies are likely to initially emerge to resolve conflicts between families. Similarly, @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military formidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Increased sedentarization and defensibility of resources is often associated with social stratification, broader institutionalized managerial roles, and stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military [@lopez_evolutionary_2017].

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. 
<!-- The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies.  -->
In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexts. 
<!-- In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations.  -->
Community leadership in WEIRD populations is likely be largely focused on economic functions, creating and managing organizational structures, and managing inter-group relations, whereas in less WEIRD contexts community leadership functions are more likely to focus on resolving interpersonal conflicts and managing intra-group dynamics.

@@

<!-- UNIVERSAL QUALITIES OF COMMUNITY LEADERS -->
There are some broad commonalities suggested in the qualities of community leaders across diverse populations. When strong community leaders emerge, they are most likely to be high status males who embody the qualities most respected by the group and necessary for successful leadership in the contexts in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 (and others) in conceptualizing social status as a relative indicator of the social value of the individual by the group and of their relative access to contested resources. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014], however, any causality between the two is likely complex, context-specific, and driven by underlying correlates.

>>>>

Research on leadership within anthropology extends to the birth of the discipline. Early on, anthropologists (and some sociologists) realized that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Anthropologists have documented the emergence of leaders among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and in diverse contexts, such as kin groups [@dussart2000politics], ritual [@singh_cultural_2017], economic groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

@@

<!-- Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and nonindustrial societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Such approaches can also incorporate the anthropological insight that diversity of cultural, social, and ecological contexts are strong drivers of human behavior and group dynamics [@wiessner_risk_1982; @henrich_big_2015]. -->

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
<!-- Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group contex. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020]. -->

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. One outstanding question for evolutionary leadership scholars is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking the nature of leadership and followership across social contexts and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. A second outstanding question is what benefits *and* costs do both leaders and followers incur. While the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [for discussions on costs of leadership see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

<!-- Leadership roles are also often associated with socially sanctioned rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].  -->

<!-- CLOSING -->
We first briefly review the qualities of leaders, the functions they serve, and the costs and benefits of leadership across nonindustrial non-WEIRD human societies. We then outline the aims and methods of the current cross-cultural study -- which focuses on exploring patterns among non-WEIRD populations -- before presenting results and discussion.

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009, drawing on the ethnography of highland New Guinea, proposes that leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling military formidability to rival groups. Among many ethnographically described small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when group demands require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Leaders also play an important role in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011] and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Representing the group and managing internal and external relationships and conflicts are expected to be the most salient functions of smaller-scale community leaders.

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Increased sedentarization and defensibility of resources is often associated with social stratification, broader institutionalized managerial roles, and stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. 
<!-- The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies.  -->
In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexts. 
<!-- In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations.  -->
Community leadership in WEIRD populations is likely be largely focused on economic functions, creating and managing organizational structures, and managing inter-group relations, whereas in less WEIRD contexts community leadership functions are more likely to focus on resolving interpersonal conflicts and managing intra-group dynamics. We include detiled review of this literature on WEIRD populations in the SI.

@@

<!-- UNIVERSAL QUALITIES OF COMMUNITY LEADERS -->
There has been increasing interdisciplinary focus on the closely related constructs of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. Whereas leaders have disproportionate group influence, status is often conceptualized as a relative indicator of the social value of the individual by the group and of their relative access to contested resources [e.g., @blader_whats_2014; @von_rueden_roots_2014]. The relationship between leadership and status is often expected to be colinear [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014]. The causal relationship between the two, however, is likely to be complex, context-specific, and driven by underlying correlates. Community leaders are frequently conceptualized as high status males who embody the qualities most respected by the group and necessary for successful leadership in the contexts in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019].
